Translation Elicitation

### 2021-05-26 Rmd/report.Rmd
Drop onset from model and fix y-axis in interacion plots.

---
title: "Translation Elicitation"
output:
  html_document:
    df_print: paged
    toc: true
    toc_depth: 2
    toc_float: true
---

@@

# import models
fit_jtrace <- readRDS(here("Results", "fit_jtrace.rds"))
fit_responses <- readRDS(here("Results", "fit_responses_3.rds"))
loo <- readRDS(here("Results", "loo.rds"))

@@

#### Vowel and consonant similarity

@@

@@

* fit_0: `~ 1 + (1  | participant)`
* fit_1: `~ pthn + (1 + pthn | participant)`
* fit_2: `~ pthn*lv + (1 + pthn*lv | participant)`

@@

```{stan stan_model, echo=TRUE, output.var="model"}
// generated with brms 2.15.0
functions {
 /* compute correlated group-level effects
  * Args: 
  *   z: matrix of unscaled group-level effects
  *   SD: vector of standard deviation parameters
  *   L: cholesky factor correlation matrix
  * Returns: 
  *   matrix of scaled group-level effects
  */ 
  matrix scale_r_cor(matrix z, vector SD, matrix L) {
    // r is stored in another dimension order than z
    return transpose(diag_pre_multiply(SD, L) * z);
  }
}
data {
  int<lower=1> N;  // total number of observations
  int Y[N];  // response variable
  int<lower=1> K;  // number of population-level effects
  matrix[N, K] X;  // population-level design matrix
  // data for group-level effects of ID 1
  int<lower=1> N_1;  // number of grouping levels
  int<lower=1> M_1;  // number of coefficients per level
  int<lower=1> J_1[N];  // grouping indicator per observation
  // group-level predictor values
  vector[N] Z_1_1;
  vector[N] Z_1_2;
  vector[N] Z_1_3;
  vector[N] Z_1_4;
  vector[N] Z_1_5;
  vector[N] Z_1_6;
  vector[N] Z_1_7;
  vector[N] Z_1_8;
  int<lower=1> NC_1;  // number of group-level correlations
  int prior_only;  // should the likelihood be ignored?
}
transformed data {
  int Kc = K - 1;
  matrix[N, Kc] Xc;  // centered version of X without an intercept
  vector[Kc] means_X;  // column means of X before centering
  for (i in 2:K) {
    means_X[i - 1] = mean(X[, i]);
    Xc[, i - 1] = X[, i] - means_X[i - 1];
  }
}
parameters {
  vector[Kc] b;  // population-level effects
  real Intercept;  // temporary intercept for centered predictors
  vector<lower=0>[M_1] sd_1;  // group-level standard deviations
  matrix[M_1, N_1] z_1;  // standardized group-level effects
  cholesky_factor_corr[M_1] L_1;  // cholesky factor of correlation matrix
}
transformed parameters {
  matrix[N_1, M_1] r_1;  // actual group-level effects
  // using vectors speeds up indexing in loops
  vector[N_1] r_1_1;
  vector[N_1] r_1_2;
  vector[N_1] r_1_3;
  vector[N_1] r_1_4;
  vector[N_1] r_1_5;
  vector[N_1] r_1_6;
  vector[N_1] r_1_7;
  vector[N_1] r_1_8;
  // compute actual group-level effects
  r_1 = scale_r_cor(z_1, sd_1, L_1);
  r_1_1 = r_1[, 1];
  r_1_2 = r_1[, 2];
  r_1_3 = r_1[, 3];
  r_1_4 = r_1[, 4];
  r_1_5 = r_1[, 5];
  r_1_6 = r_1[, 6];
  r_1_7 = r_1[, 7];
  r_1_8 = r_1[, 8];
}
model {
  // likelihood including constants
  if (!prior_only) {
    // initialize linear predictor term
    vector[N] mu = Intercept + rep_vector(0.0, N);
    for (n in 1:N) {
      // add more terms to the linear predictor
      mu[n] += r_1_1[J_1[n]] * Z_1_1[n] + r_1_2[J_1[n]] * Z_1_2[n] + r_1_3[J_1[n]] * Z_1_3[n] + r_1_4[J_1[n]] * Z_1_4[n] + r_1_5[J_1[n]] * Z_1_5[n] + r_1_6[J_1[n]] * Z_1_6[n] + r_1_7[J_1[n]] * Z_1_7[n] + r_1_8[J_1[n]] * Z_1_8[n];
    }
    target += bernoulli_logit_glm_lpmf(Y | Xc, mu, b);
  }
  // priors including constants
  target += normal_lpdf(b | 0, 3);
  target += normal_lpdf(Intercept | 0, 3);
  target += cauchy_lpdf(sd_1 | 0, 3)
    - 8 * cauchy_lccdf(0 | 0, 3);
  target += std_normal_lpdf(to_vector(z_1));
  target += lkj_corr_cholesky_lpdf(L_1 | 5);
}
generated quantities {
  // actual population-level intercept
  real b_Intercept = Intercept - dot_product(means_X, b);
  // compute group-level correlations
  corr_matrix[M_1] Cor_1 = multiply_lower_tri_self_transpose(L_1);
  vector<lower=-1,upper=1>[NC_1] cor_1;
  // extract upper diagonal of correlation matrix
  for (k in 1:M_1) {
    for (j in 1:(k - 1)) {
      cor_1[choose(k - 1, 2) + j] = Cor_1[j, k];
    }
  }
}
```

@@

ggplot(post, aes(.value, fill = .variable)) +
  facet_wrap(~.variable, scales = "free") +
  geom_vline(xintercept = 0) +
  stat_halfeye() +
  labs(x = "Value", y = "Posterior probability density", fill = "Variable") +
  theme_custom() +
  theme(
    legend.position = "none",
    plot.caption.position = "plot",
    plot.caption = element_text(hjust = 0),
    axis.text = element_text(size= 7),
    panel.grid.major.x = element_line(colour = "grey", linetype = "dotted"),
    panel.grid.major.y = element_line(colour = "grey", linetype = "dotted")
  ) +
  ggsave(here("Figures", "responses_fixed.png"))
```

@@

```{r model_comparison, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
map(loo$loos, "estimates") %>% 
  map(function(x){
    as.data.frame(x) %>% 
      rownames_to_column("ic")
  }
  ) %>% 
  bind_rows(.id = "model") %>% 
  clean_names() %>% 
  filter(ic %in% c("elpd_loo", "looic")) %>% 
  pivot_wider(names_from = ic, values_from = c("estimate", "se")) %>% 
  left_join(rownames_to_column(as.data.frame(loo$comparison, "model"))) %>% 
  mutate(formula = c(
    "...+onset+pthn:onset+lv:onset+pthn:lv:onset+(0+onset+pthn:onset+lv:onset+pthn:lv:onset|participant)",
    "...+lv+pthn:lv+(0+lv+pthn:lv|participant)",
    "...+pthn+(0+pthn|participant)",
    "~1+(1 |participant)"
  )) %>% 
  arrange(desc(model)) %>% 
  select(formula, estimate_elpd_loo, se_elpd_loo, estimate_looic, se_looic, elpd_diff, se_diff) %>%
  rename(model = formula) %>% 
  gt() %>% 
  fmt_number(2:7) %>% 
  cols_merge(c("estimate_elpd_loo", "se_elpd_loo"), pattern = "{1} ({2})") %>% 
  cols_merge(c("estimate_looic", "se_looic"), pattern = "{1} ({2})") %>% 
  cols_merge(c("elpd_diff", "se_diff"), pattern = "{1} ({2})") %>% 
  cols_label(
    model = md("**Model**"),
    estimate_elpd_loo = md("**ELPD<sub>LOO</sub> (SE)**"),
    estimate_looic = md("**LOO-IC (SE)**"),
    elpd_diff = md("**ELPD<sub>diff</sub> (SE)**")

@@

#### Marginal effects {.tabset .tab-pills}

##### By PTHN

```{r emmeans_pthn, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
nd <- expand_grid(
  pthn = seq(min(responses_coded$pthn, na.rm = TRUE), max(responses_coded$pthn, na.rm = TRUE), by = 0.1),
  onset = unique(responses_coded$onset),
  lv = c(-1, 1)
)

m <- add_fitted_draws(nd, fit_responses, n = 20, re_formula = NA) %>% 
  mutate_at(vars(lv), function(x) paste0(x, " SD")) %>% 
  mutate(onset = paste0("Onset: ", onset))

ggplot(m, aes(pthn, .value, colour = lv, fill = lv)) +
  facet_wrap(~onset) +
  # stat_lineribbon(.width = 0.95, alpha = 0.25, size = 0) +
  stat_summary(fun = mean, geom = "line", size = 1) +
  labs(x = "Phonological neighbourhood density\n(higher frequency neighbours)",
       y = "P(Correct)", colour = "Phonological similarity",
       fill = "Phonological similarity") +
  guides(colour = guide_legend(ncol = 2)) +
  theme_custom() +
  scale_y_continuous(labels = scales::percent) +
  theme(
    legend.position = "top",
    panel.grid = element_line(colour = "grey", linetype = "dotted")
  ) +
  ggsave(here("Figures", "responses_pthn.png"))

@@

##### By phonological similarity similarity

```{r emmeans_consonant, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
nd <- expand_grid(
  pthn = c(-1, 1),
  onset = unique(responses_coded$onset),
  lv = seq(
    min(responses_coded$lv, na.rm = TRUE),
    max(responses_coded$lv, na.rm = TRUE),
    by = 0.1
  )
)

m <- add_fitted_draws(nd, fit_responses, n = 20, re_formula = NA) %>% 
  mutate_at(vars(pthn), function(x) paste0(x, " SD")) %>% 
  mutate(onset = paste0("Onset: ", onset))

ggplot(m, aes(lv, .value, colour = pthn, fill = pthn)) +
  facet_wrap(~onset) +
  # stat_lineribbon(.width = 0.95, alpha = 0.25, size = 0) +
  stat_summary(fun = mean, geom = "line", size = 1) +
  labs(x = "Consonant similarity (SD)",
       y = "P(Correct)", colour = "PTHN",
       fill = "PTHN/Vowel similarity") +
  guides(colour = guide_legend(ncol = 2)) +
  theme_custom() +
  scale_y_continuous(labels = scales::percent) +
  theme(
    legend.position = "top",
    panel.grid = element_line(colour = "grey", linetype = "dotted")
  ) +
  ggsave(here("Figures", "responses_consonant.png"))

@@

##### {-}

>>>>

---
title: "Translation Elicitation"
output:
  html_document:
    df_print: paged
    toc: true
    toc_depth: 3
    toc_float: true
---

@@

# import models
fit_jtrace <- readRDS(here("Results", "fit_jtrace.rds"))
fit_responses <- readRDS(here("Results", "fit_responses_2.rds"))
loo <- readRDS(here("Results", "loo.rds"))

@@

#### Phonological similarity

@@

#### Phonological similarity and onset

```{r trials_similarity_onset, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
trials_coded %>% 
  select(trial_id, group, lv, onset) %>% 
  ggplot(aes(lv, fill = onset, colour = onset)) +
  facet_wrap(~group, ncol = 2) +
  geom_histogram() +
  labs(x = "Phonological Leveshtein distance\n(corrected for string length)", y = "# trials",
       fill = "Phonological onset", colour = "Phonological onset") +
  theme_custom() +
  theme(
    legend.position = c(0.75, 0.25),
    panel.grid.major.y = element_line(colour = "grey", linetype = "dotted")
  )
```

@@

* fit_0: `~ 1 + frequency + (1  | participant)`
* fit_1: `~ 1 + frequency + pthn + (1 + pthn | participant)`
* fit_2: `~ 1 + frequency + pthn*lv + (1 + pthn*lv | participant)`

@@

```{stan stan_model, echo=TRUE, output.var="model"}
// generated with brms 2.15.0
functions {
/* compute correlated group-level effects
* Args: 
*   z: matrix of unscaled group-level effects
*   SD: vector of standard deviation parameters
*   L: cholesky factor correlation matrix
* Returns: 
*   matrix of scaled group-level effects
*/ 
matrix scale_r_cor(matrix z, vector SD, matrix L) {
// r is stored in another dimension order than z
return transpose(diag_pre_multiply(SD, L) * z);
}
}
data {
int<lower=1> N;  // total number of observations
int Y[N];  // response variable
int<lower=1> K;  // number of population-level effects
matrix[N, K] X;  // population-level design matrix
// data for group-level effects of ID 1
int<lower=1> N_1;  // number of grouping levels
int<lower=1> M_1;  // number of coefficients per level
int<lower=1> J_1[N];  // grouping indicator per observation
// group-level predictor values
vector[N] Z_1_1;
vector[N] Z_1_2;
vector[N] Z_1_3;
vector[N] Z_1_4;
int<lower=1> NC_1;  // number of group-level correlations
int prior_only;  // should the likelihood be ignored?
}
transformed data {
int Kc = K - 1;
matrix[N, Kc] Xc;  // centered version of X without an intercept
vector[Kc] means_X;  // column means of X before centering
for (i in 2:K) {
means_X[i - 1] = mean(X[, i]);
Xc[, i - 1] = X[, i] - means_X[i - 1];
}
}
parameters {
vector[Kc] b;  // population-level effects
real Intercept;  // temporary intercept for centered predictors
vector<lower=0>[M_1] sd_1;  // group-level standard deviations
matrix[M_1, N_1] z_1;  // standardized group-level effects
cholesky_factor_corr[M_1] L_1;  // cholesky factor of correlation matrix
}
transformed parameters {
matrix[N_1, M_1] r_1;  // actual group-level effects
// using vectors speeds up indexing in loops
vector[N_1] r_1_1;
vector[N_1] r_1_2;
vector[N_1] r_1_3;
vector[N_1] r_1_4;
// compute actual group-level effects
r_1 = scale_r_cor(z_1, sd_1, L_1);
r_1_1 = r_1[, 1];
r_1_2 = r_1[, 2];
r_1_3 = r_1[, 3];
r_1_4 = r_1[, 4];
}
model {
// likelihood including constants
if (!prior_only) {
// initialize linear predictor term
vector[N] mu = Intercept + rep_vector(0.0, N);
for (n in 1:N) {
// add more terms to the linear predictor
mu[n] += r_1_1[J_1[n]] * Z_1_1[n] + r_1_2[J_1[n]] * Z_1_2[n] + r_1_3[J_1[n]] * Z_1_3[n] + r_1_4[J_1[n]] * Z_1_4[n];
}
target += bernoulli_logit_glm_lpmf(Y | Xc, mu, b);
}
// priors including constants
target += normal_lpdf(b | 0, 3);
target += normal_lpdf(Intercept | 0, 3);
target += cauchy_lpdf(sd_1 | 0, 3)
- 4 * cauchy_lccdf(0 | 0, 3);
target += std_normal_lpdf(to_vector(z_1));
target += lkj_corr_cholesky_lpdf(L_1 | 5);
}
generated quantities {
// actual population-level intercept
real b_Intercept = Intercept - dot_product(means_X, b);
// compute group-level correlations
corr_matrix[M_1] Cor_1 = multiply_lower_tri_self_transpose(L_1);
vector<lower=-1,upper=1>[NC_1] cor_1;
// extract upper diagonal of correlation matrix
for (k in 1:M_1) {
for (j in 1:(k - 1)) {
cor_1[choose(k - 1, 2) + j] = Cor_1[j, k];
}
}
}
```

@@

ggplot(post, aes(.value, .variable, fill = .variable)) +
  geom_vline(xintercept = 0) +
  stat_halfeye() +
  labs(x = "Value", y = "Posterior probability density", fill = "Variable") +
  theme_custom() +
  theme(
    legend.position = "none",
    plot.caption.position = "plot",
    plot.caption = element_text(hjust = 0),
    # axis.text = element_text(size = 7),
    panel.grid.major.x = element_line(colour = "grey", linetype = "dotted"),
    panel.grid.major.y = element_line(colour = "grey", linetype = "dotted")
  ) +
  ggsave(here("Figures", "responses_fixed.png"))
```

@@

```{r model_comparison, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE}
map(loo$loos, "estimates") %>% 
  map(function(x){
    as.data.frame(x) %>% 
      rownames_to_column("ic")
  }
  ) %>% 
  bind_rows(.id = "model") %>% 
  clean_names() %>% 
  filter(ic %in% c("elpd_loo", "looic")) %>% 
  pivot_wider(names_from = ic, values_from = c("estimate", "se")) %>% 
  left_join(rownames_to_column(as.data.frame(loo$comparison, "model"))) %>% 
  mutate(formula = c(
    "~ 1 + frequency + pthn*lv + (1 + pthn*lv | participant)",
    "~ 1 + frequency + pthn + (1 + pthn | participant)",
    "~ 1 + frequency + (1 | participant)"
  )) %>% 
  arrange(desc(model)) %>% 
  select(formula, estimate_elpd_loo, se_elpd_loo, estimate_looic, se_looic, elpd_diff, se_diff) %>%
  rename(model = formula) %>% 
  gt() %>% 
  fmt_number(2:7) %>% 
  cols_merge(c("estimate_elpd_loo", "se_elpd_loo"), pattern = "{1} ({2})") %>% 
  cols_merge(c("estimate_looic", "se_looic"), pattern = "{1} ({2})") %>% 
  cols_merge(c("elpd_diff", "se_diff"), pattern = "{1} ({2})") %>% 
  cols_label(
    model = md("**Model**"),
    estimate_elpd_loo = md("**ELPD<sub>LOO</sub> (SE)**"),
    estimate_looic = md("**LOO-IC (SE)**"),
    elpd_diff = md("**ELPD<sub>diff</sub> (SE)**")

@@

#### Marginal effects

```{r emmeans_pthn, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE, fig.width=6, fig.height=6}

# by pthn
nd <- expand_grid(
  pthn = seq(min(responses_coded$pthn, na.rm = TRUE), max(responses_coded$pthn, na.rm = TRUE), by = 0.1),
  frequency = 0,
  lv = c(-1, 1)
)

m <- add_fitted_draws(nd, fit_responses, n = 20, re_formula = NA) %>% 
  mutate_at(vars(lv), function(x) paste0(x, " SD"))

ggplot(m, aes(pthn, .value, colour = lv, fill = lv)) +
  stat_lineribbon(.width = 0.95, alpha = 0.25, size = 0) +
  stat_summary(fun = mean, geom = "line", size = 1) +
  labs(x = "PTHN", y = "P(Correct)", colour = "Phon. similarity", fill = "Phon. similarity") +
  scale_y_continuous(labels = scales::percent, limits= c(0, 1)) +
  guides(colour = guide_legend(ncol = 2)) +
  theme_custom() +
  theme(
    legend.position = "top",
    panel.grid = element_line(colour = "grey", linetype = "dotted")
  ) 
ggsave(here("Figures", "responses_pthn.png"))

@@

```{r emmeans_lv, echo=FALSE, message=FALSE, warning=FALSE, paged.print=FALSE, fig.width=6, fig.height=6}
# by lv
nd <- expand_grid(
  pthn = c(-1, 1),
  frequency = 0,
  lv = seq(
    min(responses_coded$lv, na.rm = TRUE),
    max(responses_coded$lv, na.rm = TRUE),
    by = 0.1
  )
)

m <- add_fitted_draws(nd, fit_responses, n = 20, re_formula = NA) %>% 
  mutate_at(vars(pthn), function(x) paste0(x, " SD"))

ggplot(m, aes(lv, .value, colour = pthn, fill = pthn)) +
  stat_lineribbon(.width = 0.95, alpha = 0.25, size = 0) +
  stat_summary(fun = mean, geom = "line", size = 1) +
  labs(x = "Phon. similarity", y = "P(Correct)", colour = "PTHN", fill = "PTHN") +
  scale_y_continuous(labels = scales::percent, limits= c(0, 1)) +
  guides(colour = guide_legend(ncol = 2)) +
  theme_custom() +
  theme(
    legend.position = "top",
    panel.grid = element_line(colour = "grey", linetype = "dotted")
  ) +
  ggsave(here("Figures", "responses_lv.png"))

@@
